Predicting Personal Total Income of Individuals in the United States in 2023

### 2024-11-18 paper/paper.qmd
Add alpha on the residual plots and modify the paper

---
title: "Personal Total Income of Individual in the United States of America in 2023"
subtitle: "A Predictive Model Using the American Community Survey Data"
author: 
  - Justin (Jiazhou) Bi and Weiyang Li
thanks: "Inspired and instructed by: https://github.com/RohanAlexander/marriage. All the project related files can be found at: https://github.com/Jiazhou-Bi/ACS-Income-Model/tree/main."
date: Oct 25, 2024
date-format: long
abstract: "This project aims to build a predictive model for personal total income using various demographic attributes, such as marital status, occupation, and residing state. We explored three popular machine learning algorithms: Random Forest, Linear Regression, and Extreme Gradient Boosting. Ultimately, Extreme Gradient Boosting was selected due to its superior predictive performance. While the model achieved a high R-squared value, it also exhibited a high Mean Squared Error, indicating challenges in accurately predicting all data points. Future steps for improving the model may involve incorporating additional demographic features to enhance model accuracy and reduce errors. Additionally, fine-tuning hyperparameters, experimenting with feature engineering, and handling outliers could further improve performance."

@@

Income is a fundamental concept as it affects nearly every member of society, influencing access to resources, quality of life, and overall economic stability. At the same time, it is also a highly complex concept to define, as it can be interpreted differently depending on various perspectives, such as economic, social, legal, and tax purposes. Therefore, we must examine various demographical factors that may impact income to understand its concept better. Our project defines income as "all forms of net financial resources generated or lost from work, investments, or other activities." This project aims to predict personal income using data from the 2023 American Community Survey (ACS) (@ipums_usa_v15). The ACS provides detailed demographic information, including age, education, employment, marital status, mortgage status, veteran status, etc... As a result, it provides a robust dataset for studying potential factors that may influence income.

Due to the timeframe limitations, the findings of this project are specifically applicable to the 2023 dataset. As such, the results may vary across different periods, and their generalizability has not been assessed within this project. The raw data used in this analysis was sourced from IPUMS using the IPUMS API. (@ipums_api) We utilized pandas (@pandas) for data processing, and data visualization was carried out using Seaborn. (@seaborn) Additionally, all tables presented in this report were generated with Plotly. (@plotly)

Income inequality is a serious problem and can lead to significant consequences and it may affect social structure, economic stability, as well as policy-making processes. Researchers found that demographic factors, such as education level and occupation type, can significantly impact an individual's financial standing. (@Autor_2013) Similarly, social relationships, such as marital status and household characteristics, affect income dynamics. (@WaiteGallagher_2000) Therefore, applying machine learning models to analyze the ACS dataset and using the information mentioned to predict personal income is reasonable.

We have chosen 2023 ACS dataseet because it offers many demographical details, as well as self-reported total income of the individuals. The sample is also representative and covers all geographical locations in the USA. Furthermore, it provides easy access to their database through API. We will employ advanced predictive models, including random forest, lineaer regression, and extreme gradient boosting machine to estimate personal income based on demographic factors to provide insights into which potential factors have the most significant impact on personal income. We hope the findings generated from our project can provide insights to policy-making process so that we can, hopefully, reduce income inequality and support further economic development in the future.

@@

@@

The dataset used here is downloaded from the IPUMS USA website via @ipums_usa_v15. Our project's extracted dataset contains all the responses from the American Community Survey (ACS) from 2023. Our extraction included the following variables: total income, dwelling mortgage status, sex of the individual, marital status, education attainment, private or public school, industry or occupation, veteran status, and age. We acknowledge that some of the variables (such as educational level, occupation, etc.) we have chosen here may have a more detailed version that contains more information. However, due to the nature of categorical variables, having detailed variables will create more classes of that variable, thus leading to more features for our model when encoding those variables. Although more features will improve the overall performance of our predictive model, it will also create issues such as overfitting, needing more computational resources, and lack of interpretability. Due to the trade-off of these mentioned pros and cons, we have decided to proceed with these higher-level variables for this project for simplicity. Furthermore, we have cleaned the data for better data quality. Details of our data cleaning processes are provided in @sec-datacleaning. In the following subsections, we will review all the variables used in this report and provide some basic descriptive statistics.

## Mortgage
The Mortgage variable describes the mortgage status of dwelling. There are five different values, they are "N/A", "No, owned free and clear", "Check mark on manuscript (probably yes)", "Yes, mortgaged/ deed of trust or similar debt", and "Yes, contract to purchase". We have dropped the N/A values for better data quality. Details available in @sec-datacleaning. The count of the remaining classes is shown below in (@fig-figure1).

@@

#| echo: false
#| label: fig-figure1
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Mortgage Status Distribution

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
plt.figure(figsize=(5, 3))
sns.countplot(x='MORTGAGE', data=data, palette='pastel', edgecolor='black')
plt.xlabel('Mortgage Status', fontsize=12)
plt.ylabel('Count', fontsize=12)
plt.xticks(rotation=30, ha='right')
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

@@

## Sex
This variable is coded as a binary variable, and we will use it as-is in our project. This does not suggest that the authors agree with this type of sex classification, but rather that it aligns with the available data and serves the research purposes of this project. The binary coding is a practical choice to maintain consistency with the dataset, though it may not fully capture the complexity and diversity of gender identities. @fig-figure2 below shows the distribution of this variable.

@@

#| echo: false
#| label: fig-figure2
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Sex Distribution

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
plt.figure(figsize=(5, 3))
sns.countplot(x='SEX', data=data, palette='pastel', edgecolor='black')
plt.xlabel('Sex', fontsize=12)
plt.ylabel('Count', fontsize=12)
plt.xticks(ha='right')
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

@@

## Age
Age describes the age of the individual, and a age of 0 was coded for the individuals under 1 year old. For our dataset, the minimum age is 16, however. Age distribution of our dataset is shown below in @fig-figure3.
```{python}

#| echo: false
#| label: fig-figure3
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Age Distribution

@@

## Marital Status
Marital status describes the marital status of the individual. There are 7 different values, namely "Married, spouse present", "Married, spouse absent", "Separated", "Divorced", "Widowed", "Never married/single", and "Blank, missing". We have dropped the N/A values for better data quality. Details available in @sec-datacleaning. The following @fig-figure4 illustrate the distribution of marital status of the participants in our dataset.
```{python}

#| echo: false
#| label: fig-figure4
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Counts of Marital Status

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
marrital_status_label = {
  '1': 'Married, spouse present',
  '2': 'Married, spouse absent',
  '3': 'Separated',
  '4': 'Divorced',
  '5': 'Widowed',
  '6': 'Never married/single',
  '9': 'Blank, missing'
}
data['MARST'] = data['MARST'].map(marrital_status_label)
plt.figure(figsize=(10, 6))
sns.countplot(x='MARST', data=data, palette='pastel', edgecolor='black')
plt.xlabel('Marital Status', fontsize=12)
plt.ylabel('Count', fontsize=12)
plt.xticks(rotation=30, ha='center')
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

@@

## Educational Attainment [general version]
This is a general version (compared to the detailed version). The reason of choosing the general version is mentioned in @sec-data above. This column represents the highest educational attainment achieved by the individual. The distribution of the variable is shown in @fig-figure5.
```{python}

#| echo: false
#| label: fig-figure5
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Educational Attainment

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
educ_new_label = {
    0: 'No Schooling',
    1: 'Nursery School',
    2: 'Primary School',
    3: 'Middle School',
    4: 'College 1-4 Years',
    5: 'College 5+ Years'
}
data['EDUC_new'] = data['EDUC_new'].map(educ_new_label)
plt.figure(figsize=(10, 6))
sns.countplot(x='EDUC_new', data=data, palette='pastel', edgecolor='black')
plt.xlabel('Educational Attainment', fontsize=12)
plt.ylabel('Count', fontsize=12)
plt.xticks(rotation=30, ha='center')
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

@@

## School Type
School type indicates if the individual went to a public school or private school. Other than missing values, there are three different classes. They are "Not enrolled", "Public school", and "Private school". We are not sure how the participants would respond if they have been to more than one type of schools. @fig-figure6 shows the distribution of our sample.
```{python}

#| echo: false
#| label: fig-figure6
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: School Type

@@

```
## Industry
Industry describes the type of industry the individual works in. Same as the educational attainment variable, more detailed version of this variable is available (i.e., Occupation). There are more than 200 industries listed in our sample. A detailed description can be seen at [IPUMS USA](https://usa.ipums.org/usa-action/variables/IND1990#description_section).

## Veteran Status [General Version]
Veteran status idicates whether the individual is a veteran or not. It can be either "Yes", "No", or "N/A". Please notice that "N/A" here is different from missing, suggesting this might be a different indicator, which is not explicitly explained from the ddl file provided. There is also a detailed version of this variable, which is not used in our project. The counts is displayed in @tbl-table1.
```{python}

#| echo: false
#| label: tbl-table1
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Veteran Status

@@

## Total Income
Total income is the total income made by the individual for the year. The distribution of total income is depicted in @fig-figure7. We have removed the outliers from the graph to have a better understanding of the general pattern of the overall distribution.
```{python}

#| echo: false
#| label: fig-figure7
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Total Income Distribution

@@

```
# Exploratory Data Analysis {#sec-eda}

## Income Distribution by Mortgage Status

@@

## Correlation Analysis of Age, Education, and Income

@@

## Average Income by Education and Gender {#sec-eda3}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../other/graphs/EDAinteraction.png}
    \caption{Average Income by Education and Sex}
    \label{fig:income-education-sex}
\end{figure}

@@

@@

Note that according to our previous analysis in @sec-eda3, the interaction term, `EDUC_SEX_INTERACTION`, was included to capture the combined effect of education level and gender on income. This allowed us to model the non-linear relationship between these two variables and their joint impact on the outcome variable.

@@

| Feature                  | VIF       |
|--------------------------|-----------|
| cat__STATEICP            | 3.242254  |
| cat__GQ                  | 1.002419  |
| cat__MORTGAGE            | 2.658013  |
| cat__MARST               | 1.741158  |
| cat__SCHLTYPE            | 1.239412  |
| cat__OCC2010             | 3.187731  |
| cat__IND1990             | 8.490997  |
| num__AGE                 | 7.871046  |
| num__EDUC_SEX_INTERACTION| 6.945850  |
:::

@@

| Metric                     | Value              |
|----------------------------|--------------------|
| R-squared                  | 0.0774938018       |
| Adjusted R-squared         | 0.0774637241       |
| Mean Squared Error (MSE)   | 7,448,800,507.01   |
| Root Mean Squared Error (RMSE) | 86,306.43        |
:::

@@

| Feature         | Importance |
|-----------------|------------|
| cat\_\_EDUC_new | 0.337065   |
| cat\_\_IND1990  | 0.243454   |
| num\_\_AGE      | 0.193926   |
| cat\_\_SEX      | 0.149342   |
| cat\_\_MARST    | 0.055948   |
| cat\_\_MORTGAGE | 0.013064   |
| cat\_\_SCHLTYPE | 0.004251   |
| cat\_\_VETSTAT  | 0.002948   |
:::

@@

One consistent finding across all models is the significant impact of education and industry on income. These factors were consistently ranked as the most important predictors in both Random Forest and XGBoost models. Age also played an important role but to a lesser extent, while variables such as marital status and sex showed moderate influence.

The residual analysis revealed that none of the models perfectly fit the data, particularly for higher income levels. Both Random Forest and XGBoost exhibited a pattern of increasing residual variance (heteroscedasticity) as predicted income increased, which suggests that the models struggled to accurately predict high-income values. Linear Regression, while more interpretable, performed the worst in terms of accuracy, especially for non-linear relationships between the predictors and income.

Despite the overall success of XGBoost, it is clear that there are areas where the model's performance could be further refined.

>>>>

---
title: "Predicting Personal Total Income of Individuals in the United States in 2023"
subtitle: "XGBoost Captures the Complex Pattern Better Than Random Forest and Linear Regression"
author: 
  - Justin (Jiazhou) Bi and Weiyang Li
thanks: "All project related files available at: https://github.com/Jiazhou-Bi/ACS-Income-Model"
date: Nov 14, 2024
date-format: long
abstract: "This project aims to build a predictive model for personal total income using various demographic attributes, such as marital status, occupation, and residing state. We explored three popular machine learning algorithms: Random Forest, Linear Regression, and Extreme Gradient Boosting. Ultimately, Extreme Gradient Boosting was selected due to its superior predictive performance. While the model achieved a high R-squared value, it also exhibited a high Mean Squared Error, indicating challenges in accurately predicting all data points. Future steps for improving the model may involve incorporating additional demographic features to enhance model accuracy and reduce errors. Additionally, fine-tuning hyperparameters, experimenting with feature engineering, and handling outliers could further improve performance."

@@

Income is a fundamental concept as it affects nearly every member of society, influencing access to resources, quality of life, and overall economic stability. At the same time, it is also a difficult concept to define, as it can be interpreted differently depending on various perspectives, such as economic, social, legal, and tax purposes. Therefore, we must examine various demographical factors that may impact income to understand its concept better. Our project defines income as "all forms of net financial resources generated or lost from work, investments, or other activities." This project aims to predict personal income using data from the 2023 American Community Survey (ACS) [@ipums_usa_v15]. The ACS provides detailed demographic information, including age, education, employment, marital status, mortgage status, veteran status. As a result, it provides a robust dataset for studying potential factors that may influence income.

Due to the timeframe limitations, the findings of this project are specifically applicable to the 2023 dataset. As such, the results may vary across different periods, and their generalizability has not been assessed within this project. The raw data used in this analysis was sourced from IPUMS using the IPUMS API [@ipums_api]. We utilized pandas [@pandas] for data processing, and data visualization was carried out using Seaborn [@seaborn]. Additionally, all tables presented in this report were generated with Plotly [@plotly].

Income inequality is a serious problem and can lead to significant consequences and it may affect social structure, economic stability, as well as policy-making processes. Researchers found that demographic factors, such as education level and occupation type, can significantly impact an individual's financial standing [@Autor_2013]. Similarly, social relationships, such as marital status and household characteristics, affect income dynamics [@WaiteGallagher_2000]. Therefore, applying machine learning models to analyze the ACS dataset and using the information mentioned to predict personal income is reasonable.

We have chosen 2023 ACS data because it offers many demographical details, as well as self-reported total income of the individuals. The sample is also representative and covers all geographical locations in the USA. Furthermore, it provides easy access to their database through API. We use predictive models, including random forest, linear regression, and extreme gradient boosting machine to estimate personal income based on demographic factors to provide insights into which potential factors have the most significant impact on personal income. We hope the findings generated from our project can provide insights to policy-making process so that we can, hopefully, reduce income inequality and support further economic development in the future.

@@

In this analysis, we explored three different models—Linear Regression, Random Forest, and XGBoost—to predict income based on various demographic factors. Each model provides valuable insights into the relationships between income and predictors, such as education, age, industry, and gender. Among the models, XGBoost demonstrated the best performance in terms of predictive accuracy, as evidenced by its lower RMSE and superior handling of complex data interactions.

The remainder of this paper is structured as follows: @sec-data introduces the raw dataset and describes the cleaned datasets, along with a preliminary analysis through numerical summaries and visualizations. @sec-model describes the three machine learning models being used in our study. @sec-results explores key findings from our analysis. Lastly, @sec-discussion addresses the limitations of the analysis and offers recommendations for reducing delays and improving subway service reliability.

@@

The dataset used here is downloaded from the IPUMS USA website via [@ipums_usa_v15]. Our project's extracted dataset contains all the responses from the American Community Survey (ACS) from 2023. Our extraction included the following variables: total income, dwelling mortgage status, sex of the individual, marital status, education attainment, private or public school, industry or occupation, veteran status, and age. Some of the variables (such as educational level, occupation) we have chosen here have a more detailed version that contains more information. However, due to the nature of categorical variables, having detailed variables will create more classes of that variable, thus leading to more features for our model when encoding those variables. Although more features will improve the overall performance of our predictive model, it will also create issues such as overfitting, needing more computational resources, and lack of interpretability. We have chosen higher-level variables in this study for better computational efficiency. Furthermore, we have cleaned the data for better data quality. Details of our data cleaning processes are provided in @sec-datacleaning. In the following subsections, we will review all the variables used in this report and provide some basic descriptive statistics.

## Descriptive Data Analysis

In this subsection, we will explore every variable used in our study by examining some of the basic statistics. These preliminary analyses can help us better understand the data we are working with.

### Mortgage
The Mortgage variable describes the mortgage status of dwelling. There are five different values, they are "N/A", "No, owned free and clear", "Check mark on manuscript (probably yes)", "Yes, mortgaged/ deed of trust or similar debt", and "Yes, contract to purchase". We have dropped the N/A values for better data quality. Details available in @sec-datacleaning. The count of the remaining classes is shown below in (@fig-figure1).

@@

#| echo: false
#| label: fig-figure1
#| warning: false
#| message: false
#| fig-pos: "htp"
#| fig-cap: Most Homeowners Are on Mortgage

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
mortgage_labels = {
    '1': 'No, owned free and clear',
    '3': 'Yes, mortgaged',
    '4': 'Yes, contract to purchase'
}
data['MORTGAGE'] = data['MORTGAGE'].astype(str).map(mortgage_labels)
plt.figure(figsize=(5, 3))
sns.countplot(x='MORTGAGE', data=data, palette='pastel', edgecolor='black')
plt.xlabel('Mortgage Status', fontsize=12)
plt.ylabel('Count', fontsize=12)
plt.xticks(rotation=30, ha='right')
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

@@

### Sex
This variable is coded as a binary variable, and we will use it as-is in our project. This does not suggest that the authors agree with this type of sex classification, but rather that it aligns with the available data and serves the research purposes of this project. The binary coding is a practical choice to maintain consistency with the dataset, though it may not fully capture the complexity and diversity of gender identities. @tbl-table1 shows the distribution of this variable.

@@

#| echo: false
#| label: tbl-table1
#| warning: false
#| message: false
#| fig-pos: "htp"

data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
sex_labels = {'1': 'Male', '2': 'Female'}
data['SEX'] = data['SEX'].map(sex_labels)
vet_status_count = data['SEX'].value_counts().reset_index()
vet_status_count.columns = ['Sex', 'Count']
fig, ax = plt.subplots(figsize=(6, 2))
ax.axis('off')
table = ax.table(
    cellText=vet_status_count.values,
    colLabels=vet_status_count.columns,
    cellLoc='center',
    loc='center'
)
table.auto_set_font_size(False)
table.set_fontsize(10)
table.scale(1, 1.5)
plt.tight_layout()
plt.show()

@@

### Age
Age describes the age of the individual, and a age of 0 was coded for the individuals under 1 year old. For our dataset, the minimum age is 16, however. Age distribution of our dataset is shown below in @fig-figure2.
```{python}

#| echo: false
#| label: fig-figure2
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Age Distribution

@@

### Marital Status
Marital status describes the marital status of the individual. There are 7 different values, namely "Married, spouse present", "Married, spouse absent", "Separated", "Divorced", "Widowed", "Never married/single", and "Blank, missing". We have dropped the N/A values for better data quality. Details available in @sec-datacleaning. The following @fig-figure3 illustrate the distribution of marital status of the participants in our dataset.
```{python}

#| echo: false
#| label: fig-figure3
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Counts of Marital Status, Most Participants Are Married with Spouse Present, or Never Married/Single

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
marrital_status_label = {
  '1': 'Married, spouse present',
  '2': 'Married, spouse absent',
  '3': 'Separated',
  '4': 'Divorced',
  '5': 'Widowed',
  '6': 'Never married/single',
  '9': 'Blank, missing'
}
order = [
    'Married, spouse present',
    'Married, spouse absent',
    'Separated',
    'Divorced',
    'Widowed',
    'Never married/single'
]
data['MARST'] = data['MARST'].map(marrital_status_label)
plt.figure(figsize=(10, 6))
sns.countplot(x='MARST', data=data, order=order, palette='pastel', edgecolor='black')
plt.xlabel('Marital Status', fontsize=12)
plt.ylabel('Count', fontsize=12)
plt.xticks(rotation=30, ha='center')
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

@@

### Educational Attainment
This is a general version (compared to the detailed version). The reason of choosing the general version is mentioned in @sec-data above. This column represents the highest educational attainment achieved by the individual. The distribution of the variable is shown in @fig-figure4.
```{python}

#| echo: false
#| label: fig-figure4
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Most Participants Have Middel School Attainment or Above

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
educ_new_label = {
    0: 'No Schooling',
    1: 'Nursery School',
    2: 'Primary School',
    3: 'Middle School',
    4: 'College 1-4 Years',
    5: 'College 5+ Years'
}
order = [
    'No Schooling',
    'Nursery School',
    'Primary School',
    'Middle School',
    'College 1-4 Years',
    'College 5+ Years'
]
data['EDUC_new'] = data['EDUC_new'].map(educ_new_label)
plt.figure(figsize=(10, 6))
sns.countplot(x='EDUC_new', data=data, order=order, palette='pastel', edgecolor='black')
plt.xlabel('Educational Attainment', fontsize=12)
plt.ylabel('Count', fontsize=12)
plt.xticks(rotation=30, ha='center')
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

@@

### School Type
School type indicates if the individual went to a public school or private school. Other than missing values, there are three different classes. They are "Not enrolled", "Public school", and "Private school". We are not sure how the participants would respond if they have been to more than one type of schools. @fig-figure5 shows the distribution of our sample.
```{python}

#| echo: false
#| label: fig-figure5
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: School Type

@@

### Industry
Industry describes the type of industry the individual works in. Same as the educational attainment variable, more detailed version of this variable is available (i.e., Occupation). There are more than 200 industries listed in our sample. A detailed description can be seen at [IPUMS USA](https://usa.ipums.org/usa-action/variables/IND1990#description_section).

### Veteran Status
Veteran status idicates whether the individual is a veteran or not. It can be either "Yes", "No", or "N/A". Please notice that "N/A" here is different from missing, suggesting this might be a different indicator, which is not explicitly explained from the ddl file provided. There is also a detailed version of this variable, which is not used in our project. The counts is displayed in @tbl-table2.
```{python}

#| echo: false
#| label: tbl-table2
#| warning: false
#| message: false
#| fig-pos: "H"

@@

### Total Income
Total income is the total income made by the individual for the year. The distribution of total income is depicted in @fig-figure6. We have removed the outliers from the graph to have a better understanding of the general pattern of the overall distribution. From the figure, it is clear that more income are reported near multiple of $10,000 as a unit. This may be caused by the "heaping and leaping" phenomonon found in self-reported survey data [@pudney2008heaping].
```{python}

#| echo: false
#| label: fig-figure6
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Total Income Distribution

@@

## Analysis of Feature Relationships {#sec-afr}

In this subsection, we will explore some of the potential relationships among the features chosen for our study. Understanding these relationships can prepare us better for our model construction in the next section.

### Income Distribution by Mortgage Status

@@

### Correlation Analysis of Age, Education, and Income

@@

### Average Income by Education and Gender {#sec-afr3}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../other/graphs/EDAinteraction.png}
    \caption{Average Income by Education and Sex}
    \label{fig:income-education-sex}
\end{figure}

@@

## Train Test Split of Data

We have cleaned our data before we split it into training and testing datasets. We are aware of potential data leakage of our testing data, and we have mitigated this risk by not using overall statistics or normalization that requires the original data to calculate, not imputing any missing data, or engineering features that include the original dataset. By doing so, we can minimize the risk of data leakage, and ensure the perfect seventy and thirty percent train test split.

@@

Note that according to our previous analysis in @sec-afr3, the interaction term, `EDUC_SEX_INTERACTION`, was included to capture the combined effect of education level and gender on income. This allowed us to model the non-linear relationship between these two variables and their joint impact on the outcome variable.

@@

| Feature                  | VIF       |
|--------------------------|-----------|
| cat__STATEICP            | 3.24  |
| cat__GQ                  | 1.00  |
| cat__MORTGAGE            | 2.66  |
| cat__MARST               | 1.74  |
| cat__SCHLTYPE            | 1.24  |
| cat__OCC2010             | 3.19  |
| cat__IND1990             | 8.49  |
| num__AGE                 | 7.87  |
| num__EDUC_SEX_INTERACTION| 6.95  |
:::

@@

| Metric                     | Value              |
|----------------------------|--------------------|
| R-squared                  | 0.0774       |
| Adjusted R-squared         | 0.0774     |
| Mean Squared Error (MSE)   | 7,448,800,507.01   |
| Root Mean Squared Error (RMSE) | 86,306.43        |
:::

@@

| Feature         | Importance |
|-----------------|------------|
| cat\_\_EDUC_new | 0.337   |
| cat\_\_IND1990  | 0.243   |
| num\_\_AGE      | 0.194   |
| cat\_\_SEX      | 0.149   |
| cat\_\_MARST    | 0.056   |
| cat\_\_MORTGAGE | 0.013   |
| cat\_\_SCHLTYPE | 0.004   |
| cat\_\_VETSTAT  | 0.003   |
:::

@@

### Feature Importance
We have found that educational attainment and occupation industry are consistently the most important features in all three models, especially in XGBoost and Random Forest. This indicates that higher educational attainment and certain industries are correlated with higher personal income. This confirms our original hypothesis that certain demographic factors, such as education, may lead to specialized industry sectors where more income can be generated.

Age is also a useful predictor. However, despite the common belief that age is a determining factor when predicting personal income, its effectiveness is not as strong as education or industry. This may be caused by the fact that in some industries, your experience may hit a plateau, after which age may not contribute as much to experience as before certain age.

Marital status and sex do affect personal income, but not as significantly as the other features. It is great to prove that certain demographical features are not as informative when predicting personal income, suggesting a "fair" distribution of wealth.

### Models

None of the three models used in our study provided an excellent prediction of personal income. For the residual analysis, it appears that the models' predictbility deterioate as the self-reported income gets higher. REsidual variance increased as the predicted income increased in the Random Forest and XGBoost model, suggesting potential heteroscedasticity. Our educated guess is that extra-high income may come from various sources, and simplifying them into a simple model like this is not a reasonable solution.

Among the three models, XGBoost has the highest accuracy overall. We believe that it can capture more complex relationships and interactions among the variables than the other two models. However, the high MSE for extra-high income individuals indicates that the model still needs to be refined more to achieve an accurate prediction.

Linear Regression has underperformed relative to the other two models. We speculate that simplicity in modeling linear relationships did not provide sufficient information to capture the non-linear patterns underlying our data, especially where multiple features interact to influence income, as suggested in @sec-afr.
